---
title: "Maximum Likelihood"
author: "Eric Scott"
date: "2020-01-30"
output:
  powerpoint_presentation:
    reference_doc: "template.pptx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
```
```{r include=FALSE}
ps <- seq(0.0001, 0.9999, 0.0001)
fx <- dbinom(x = 10, size = 30, prob = ps, log = TRUE)
fprimex <- c(diff(fx) / diff(ps), NA)
fprimeprimex <- c(diff(fprimex) / diff(ps), NA)

df <- data.frame(ps, fx, fprimex, fprimeprimex)


likprof <- ggplot(df, aes(x = ps, y = fx)) +
  geom_line(color = "darkgreen") +
  ylim(-10, 0) +
  labs(x = "p", y = "ln-likelihood")

likderiv <- ggplot(df, aes(x = ps, y = fprimex)) +
  geom_hline(aes(yintercept = 0), color = "grey40") +
  geom_line(color = "blue") +
  ylim(-200, 200) +
  labs(x = "p", y = "d[ln-likelihood] / d[p]")
```

## Maximum Likelihood Estimates

```{r echo=FALSE}
likprof + ggtitle("Binomial log-likelihood for N = 30, k = 10")
```
- A **maximum likelihood estimate** (MLE) is an estimate of a parameter that maximizes likelihood.
- In this example, there is only one free parameter, $p$. 
- The maximum value of $p$ is where the slope (AKA derivative) is zero.

## Finding Maximum Likelihood

To find the maximum likelihood estimate (MLE) of our parameters...

1. Likelihood Profile (grid search of all parameter combinations)
    - slow
    - you might miss a sharp maximum
2. Calculus
    - not always solvable    
3. Clever search algorithms (R uses these)
    - assumes likelihood surface is smooth

## Calculus

- Last time we saw that for binomial distributions, $p = k/N$ (because calculus!)
- What is the maximum likelihood estimate of $p$ for 10 events and 30 trials?
- Most of the time, our models will be more complex with more than one parameter to estimate, and our likelihood profile (or surface) won't be easy to do calculus on.

## In R using clever algorithms

- `glm()` is one function that we can use to get a MLE for $p$.

- the `glm()` function fits a **generalized linear model**, a concept we will explain in more depth later in the course.

- For now, we're going to use it to get the likelihood for a "null" model (i.e. the usual shape of the distribution without any additional sources of variation).


## R code for MLE using glm()

```{r}
success <- 10
failure <- 30 - success
mydat <- cbind(success, failure)  # successes and failures, i.e., k and N-k
glm(mydat ~ 1, family = binomial(link = "identity")) # function  to estimate parameters
m1 <- glm(mydat ~ 1, family = binomial(link = "identity")) # need to save the analysis as an object to see results
logLik(m1) # extract the log likelihood, evaluated at MLE
coef(m1) # extract the maximum likelihood estimates of the coefficients
```

## Under the hood of glm()

The `glm()` function in turn calls `glm.fit()` which in turn calls `glm.control()`.

It uses an algorithm to find the parameter values that maximize likelihood. This might seem like overkill now (remember, for a simple binomial model, the MLE of $p$ is $k/N$), but a lot easier than calculating by hand for more complicated models with more parameters.

- You don't need to know how to do these calculations, R does them for you!
- So why am I telling you about this? Because common error messages encounted when running `glm()` can only be understood if you have an idea of what is going on under the hood

Terms you want to understand:

- "convergence"
- "starting values"
- "iterations"
- "tolerance"

## Maximum Likelihood algorithms

GOAL: find place where slope (derivative) = 0 in situations where you can’t solve the general equation using calculus.

```{r echo=FALSE}
likprof + ggtitle("Binomial log-likelihood for N = 30, k = 10")
```

## Newton's Method
We'll come back to maximum likelihood in a bit...  
Newton's method is a way to find the x-intercept of arbitrary functions

In a nutshel...

- Start with a guess, $x_0$
- Calculate $f(x_0)$ and the derivative (slope) at $x_0$, $f^\prime(x_0)$
- Pick a new guess, based on where the tangent line crosses the x-axis: $x_1 = x_0 - f(x_0)/f^\prime(x_0)$
- Keep going until $x_0 \approx x_1$, with some tolerance

[Graphical explanation on board]

Start with simple exponential curve.

Pick $x_0$

Draw tangent

Label $x_1$

How can we find $x_1$ mathematically?

Draw triangle.  Height of triangle = $f(x_0)$, AKA y-value at $x_0$ Base of triangle = $x_0 - x_1$.  Slope of triangle = rise/run (alternatively, $tan \theta = opposite/adjacent$).  Slope is also the derivative of the curve evaluated at $x_0$, $f^\prime(x_0)$

$$
f^\prime(x_0) = \frac{f(x_0)}{x_0 - x_1}
$$

rearrange

$$
{x_0 - x_1} = \frac{f(x_0)}{f^\prime(x_0)}
$$

$$
x_1 = x_0 - \frac{f(x_0)}{f^\prime(x_0)}
$$

### But what if I don't know how to take the derivative?

Don't worry! An approximation is good enough! R uses some default tiny $\Delta x$ and caluculates rise over run for that tiny distance to estimate the slope.

## Iterate

- After you find $x_1$, repeat to find $x_2$

$$
x_2 = x_1 - \frac{f(x_1)}{f^\prime(x_1)}
$$

A more general form:

$$
x_{n+1} = x_n - \frac{f(x_n)}{f^\prime(x_n)}
$$

Keep iterating until $x_{n+1} \approx x_n$ within some small tolerance (say, 10^-6^)

## Using Newton's Method for MLE

But we want to find a *maximum* not a *root*! We can use deriviatives to find where the *slope*, $f^\prime(x) = 0$ instead of finding where $f(x) = 0$

```{r}
likderiv + ggtitle("Deriviative of binomial log-likelihood N = 30, k = 10")
```

Alternatively, just re-write Newton's method using first derivative and second derivative.

$$
x_1 = x_0 - f^\prime(x_0)/f^{\prime \prime}(x_0)
$$

## Example

10 trials, 30 events.  What is the MLE for p(event)?

1. Pick a starting value, $p_0$
2. find p1
3. repeat until xn = xn+1

(use elizabeth's plots)



```{r include=FALSE}
p1 <- 0.1
df %>% filter(ps == 0.1) %>% mutate(p2 = ps - fprimex/fprimeprimex)
df %>% filter(ps == 0.17600) %>% mutate(p3 = ps - fprimex/fprimeprimex)
df %>% filter(ps == 0.268) %>% mutate(p4 = ps-fprimex/fprimeprimex)
df %>% filter(ps == 0.325) %>% mutate(p5 = ps-fprimex/fprimeprimex)
df %>% filter(ps == 0.333) %>% mutate(p6 = ps-fprimex/fprimeprimex)

p = c(0.1, 0.176, 0.268, 0.325, 0.333, 0.333)
fx = c(-7.914854, -4.026201, -2.188971, -1.881945, -1.877225)
```

## step 1

```{r echo=FALSE}
likprof +
  geom_segment(x = 0.1, xend = 0.176, y = fx[1], yend = fx[2], color = "purple") +
  geom_point(x = p[1], y = fx[1]) +
  geom_point(x = p[2], y = fx[2])
```
 
 
|  p_i| f'(x)|  f''(x)| p_i+}|
|--:|--:|--:|--:|
|  0.1|  77.727|  -1022.7| 0.176|


## step 2

```{r echo=FALSE}
likprof +
  geom_segment(x = 0.1, xend = 0.176, y = fx[1], yend = fx[2], color = "purple") +
  geom_point(x = p[1], y = fx[1]) +
  geom_point(x = p[2], y = fx[2]) +
    geom_segment(x = p[2], xend = p[3], y = fx[2], yend = fx[3], color = "purple") +
  geom_point(x = p[3], y = fx[3])

```
 
 
|  p_i| f'(x)|  f''(x)| p_i+}|
|--:|--:|--:|--:|
|  0.1|  77.7|  -1022.7| 0.176|
|  0.176| 32.5| -351.9| 0.268|

## Step 3

```{r echo=FALSE}
likprof +
  geom_segment(x = 0.1, xend = 0.176, y = fx[1], yend = fx[2], color = "purple") +
  geom_point(x = p[1], y = fx[1]) +
  geom_point(x = p[2], y = fx[2]) +
    geom_segment(x = p[2], xend = p[3], y = fx[2], yend = fx[3], color = "purple") +
  geom_point(x = p[3], y = fx[3]) +
  geom_segment(x = p[3], xend = p[4], y = fx[3], yend = fx[4], color = "purple") +
  geom_point(x = p[4], y = fx[4])

```
 
 
|  p_i| f'(x)|  f''(x)| p_i+}|
|--:|--:|--:|--:|
|  0.1|  77.7|  -1022.7| 0.176|
|  0.176| 32.5| -351.9| 0.268|
| 0.268| 10.0| -176.5| 0.325|

## Step 5

```{r echo=FALSE}
likprof +
  geom_segment(x = 0.1, xend = 0.176, y = fx[1], yend = fx[2], color = "purple") +
  geom_point(x = p[1], y = fx[1]) +
  geom_point(x = p[2], y = fx[2]) +
    geom_segment(x = p[2], xend = p[3], y = fx[2], yend = fx[3], color = "purple") +
  geom_point(x = p[3], y = fx[3]) +
  geom_segment(x = p[3], xend = p[4], y = fx[3], yend = fx[4], color = "purple") +
  geom_point(x = p[4], y = fx[4]) +
  geom_segment(x = p[4], xend = p[5], y = fx[4], yend = fx[5], color = "purple") +
  geom_point(x = p[5], y = fx[5])

```
 
 
|  p_i| f'(x)|  f''(x)| p_i+}|
|--:|--:|--:|--:|
|  0.1|  77.7|  -1022.7| 0.176|
|  0.176| 32.5| -351.9| 0.268|
| 0.268| 10.0| -176.5| 0.325|
| 0.325| 1.1| -138.5| 0.333|

## Step 6

```{r echo=FALSE}
likprof +
  geom_segment(x = 0.1, xend = 0.176, y = fx[1], yend = fx[2], color = "purple") +
  geom_point(x = p[1], y = fx[1]) +
  geom_point(x = p[2], y = fx[2]) +
    geom_segment(x = p[2], xend = p[3], y = fx[2], yend = fx[3], color = "purple") +
  geom_point(x = p[3], y = fx[3]) +
  geom_segment(x = p[3], xend = p[4], y = fx[3], yend = fx[4], color = "purple") +
  geom_point(x = p[4], y = fx[4]) +
  geom_segment(x = p[4], xend = p[5], y = fx[4], yend = fx[5], color = "purple") +
  geom_point(x = p[5], y = fx[5]) +
  geom_segment(x = p[5], xend = p[6], y = fx[5], yend = fx[6], color = "purple") +
  geom_point(x = p[6], y = fx[6])

```

|  p_i| f'(x)|  f''(x)| p_i+}|
|--:|--:|--:|--:|
|  0.1|  77.7|  -1022.7| 0.176|
|  0.176| 32.5| -351.9| 0.268|
| 0.268| 10.0| -176.5| 0.325|
| 0.325| 1.1| -138.5| 0.333|
| 0.333| 0.038| -135.1| 0.333|

(Convergence!!)

## Vocab Recap

- Starting values = initial guess for parameters, e.g. $p_0$
- Iterations = how many times do you need to run through the algorithm to reach convergence
- Convergence = when $p_{n+1} \approx p_n$
- Tolerance = How close to they have to be to be consider "converged"

## Real error messages you might see:

```
Warning: glm.fit: algorithm did not converge
```
OR

```
Warning: glm.fit: algorithm stopped at boundary value
```

These are not errors.  That is, `glm()` sill runs and spits out an answer.  But you should not trust the MLE of any parameters in these cases.

How to fix?  You can tell R to try harder by doing more iterations and sometimes that works!

```
Error in eval(family$initialize) : cannot find valid starting values: please specify some
```

This is an error.  That is, `glm()` doesn't run and doesn't produce any results.

How to fix?  You can supply some starting values, sometimes you can use the coeficients from a simpler model, or just take a guess.  HOWEVER, this error often results from other more complicated underlying issues.


```{r}
success <- c(0,0,0,0,0)
failure <- 30 - success

Y = cbind(success, failure)

glm(Y ~ 1, family = binomial(link = "identity"))
```

Here, it can't find starting values because it's really hard to estimate the probability of an event that never happens!  You've got bigger problems than valid starting values for `glm.fit()`!  You could try providing a starting value, but then you get siome warnings.

```{r}
glm(Y ~ 1, family = binomial(link = "identity"), start = 0.00000000001)
```


## Exercise: use glm() to get MLE for real data
oak example.  For now, get MLE for species separately--next time, learn how to combine into one model (I think)


