---
title: "Population Dynamics"
author: "Eric Scott"
date: "2/17/2020"
output:
  powerpoint_presentation:
    reference_doc: template.pptx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error = TRUE)
```
```{r include=FALSE}
library(tidyverse)
library(here)
source(here("plot_theme.R"))
library(patchwork)
library(scales)
```

## Homework 4 Feedback

1. Can't compare p-values of models to judge relative support

2. Confidence intervals
  - They don't have to be symmetrical (often aren't with link functions), so report lower and upper not ±.
  - Error bars plotted on graphs aren't always 95% CIs---need to specify in figure captions
  

# Hacking GLMs to do population growth models

## Grey Wolves

History

```{r message=FALSE, warning=FALSE}
wolves <- read_csv(here("data", "NRMwolves.csv"))
head(wolves)
wolves <- wolves %>%
  mutate(year_post = year - 1982,
         year = as.integer(year)) 

head(wolves)
```

```{r echo=FALSE}
wolfplot <-
  ggplot(wolves, aes(x = year, y = num.wolves)) +
  geom_point() 
wolfplot
```

## Linear vs. exponential growth

We could try to fit a straigt line to the wolf data, but it is going to be difficult.

$$
N_T = \beta_0 + \beta_1\times T
$$
Where $T$ is the year 1990 and $N_T$ is the number of wolves at year T

We could fit this using an "identity" link Poisson GLM...

```{r}
m1 <- glm(num.wolves ~ year, family = poisson(link = "identity"), data = wolves)
```

...but, it errors.

- Why might R be encountering this error?

## 10 years of data

Let's use a 10 year portion of this data from 1995--2005 that looks a little more reasonable as a line to demonstrate first.

```{r}
wolves_sub <- wolves %>% filter(year >= 1995 & year <= 2005)
```
```{r echo=FALSE}
wp <- ggplot(wolves_sub, aes(x = year, y = num.wolves)) +
  geom_point() + scale_x_continuous(breaks = pretty_breaks())
wp
```

## Use an identity link to fit a straight line

```{r}
m1 <- glm(num.wolves ~ year_post, family = poisson(link = "identity"), data = wolves_sub)
coef(m1)
```

- Essentially impossible.  Predicts -1002 wolves in 1982!

```{r echo=FALSE}
pred_data <- tibble(year = wolves_sub$year, num.wolves = predict(m1, type = "response"))

wp2 <-
  wp +
  geom_line(data = pred_data, color = "blue")
wp2
```

## Exponential growth

- Most populations do not grow linearly because reproduction is a *rate* that depends on current population.
- Exponential growth describes unregulated reproduction

$$
N_{t+1} = \lambda N_t
$$

where $\lambda$ is the population growth rate.

$$
\lambda = 1+births-deaths
$$

If there is no population growth (births = deaths), $\lambda = 1$


## Exponential growth

Why exponential?

If you want to know growth at time = t + 2...

$$
N_{t + 2} = \lambda N_{t + 1}
$$

$$
= \lambda\lambda N_t
$$

$$
= \lambda^2N_t
$$

More generally...

$$
N_{t+T} = \lambda^TN_t
$$

Or if you start at year 0, then it's just

$$
N_T = \lambda^TN_0
$$

So N at year 50 = 

$$
N_{50} = \lambda^{50}N_0
$$

## Exponential growth

If lambda < 1

```{r fig.width=7}
births = 0.05
deaths = 0.15
lambda = 1 + births - deaths
N0 = 50 #population size at year 0

fake_pop <- tibble(
  year = 0:50,
  N = N0 * lambda ^ year
)
ggplot(fake_pop, aes(x = year, y = N)) +
  geom_line() +
  labs(subtitle = paste("lambda =", lambda)) +
  
ggplot(fake_pop, aes(x = year, y = log(N))) +
  geom_line()
#the patchwork package allows you to make two-panel plots like this
```

## Exponential growth

When lambda > 1

```{r fig.width=7}
births = 0.15
deaths = 0.05
lambda = 1 + births - deaths
N0 = 50 #population size at year 0

fake_pop <- tibble(
  year = 0:50,
  N = N0 * lambda ^ year
)

ggplot(fake_pop, aes(x = year, y = N)) +
  geom_line() +
  labs(subtitle = paste("lambda =", lambda)) +

ggplot(fake_pop, aes(x = year, y = log(N))) +
  geom_line()
#the patchwork package allows you to make two-panel plots like this
```

## Use log-link instead
A line with a log-link is an exponential curve when back-transformed!

$$
log(y) = \beta_0 + \beta_1x_1
$$

$$
y = e^{\beta_0} \times e^{\beta_1x_1}
$$
Our model more specifically:

$$
ln(N_T) = \beta_0 + \beta_1T
$$

So:

$$
N_T = e^{\beta_0} \times e^{\beta_1 T}
$$

$$
N_T = e^{\beta_0} \times (e^{\beta_1}) ^T
$$

Compare with exponential growth equation

## Fit exponential growth model

We can use a log-link to fit an exponential growth model to the wolf data.

```{r}
m_exp <- glm(num.wolves ~ year_post, 
             family = poisson(link = "log"), data = wolves)
exp(coef(m_exp))
```

$$
N_T = 19.615 \times1.177^T
$$

- The backtransformed coefficient for `year_post` = $\lambda$, the population growth rate
- The backtransformed intercept is the estimate for $N_0$, the number of wolves at `year_post` = 0

```{r echo=FALSE}
wolfplot +
  geom_line(aes(y = predict(m_exp, type = "response")), color = "darkgreen")
```

Hmm... doesn't look quite right.

## Density-dependent growth

Exponential growth isn't realistic over the long term for most populations, even under favorable conditions.

Density dependent factors:

  - limited space
  - limited food
  - faster disease spread
  - agression
  - emigration
  - etc.
  
These things limit exponential growth.  We can model this a couple of ways.


## Observation error vs. process error

Observation error:

- year to year variation (from predicted results) is due to imperfect estimation of wolves (undercount some years, overcount others).

Process error:

- Population growth rate depends on population size
- We can model process error by estimating $N_{t+1}/N_{t}$ (a rate).

## Process error using offset

We used offsets previously to estimate rates of bee visitation (bees/hour)

Here, we "hack" a glm to use offsets to fit a process error model.  That is, we want to model a rate of increase relative to the previous year.

$$
log(N_{t+1}/N_{t}) = \beta_0
$$

$$
log(N_{t+1}) = \beta_0 + log(N_{t})
$$

## Lagged data

We can create a column of the number of wolves in the previous year with the `lag()` function.

```{r}
wolves2 <- wolves %>%
  mutate(num.prev = lag(num.wolves)) %>%
  select(year, year_post, num.wolves, num.prev)
head(wolves2)
```

We want to remove the first year, where we don't have any data on the previous year's number of wolves.  `is.na(x)` returns `TRUE` if something is `NA`, but we want the opposite of that, so we negate it with `!`

```{r}
wolves2 <- wolves2 %>% filter(!is.na(num.prev))
head(wolves2)
```

## Fit process error model

```{r}
m_process <- glm(num.wolves ~ 1, offset = log(num.prev),
                 family = poisson(link = "log"), data = wolves2)

exp(coef(m_process))
```

This is the yearly rate of increase (on a log scale).

$$
N_{t+1} = e^{\beta_0} \times N_{t}
$$
So, if there are 13 wolves in 1985, how many would it predict in 1986?

```{r}
1.108238 * 13
```

```{r echo=FALSE}
ggplot(wolves2, aes(x = year, y = num.wolves)) +
  geom_point() +
  geom_line(aes(y = predict(m_process, type = "response")), color = "orange")
```

## Ricker model

A Ricker model takes carrying capacity into account and allows growth rate to change as the population increases.

$$
N_{t+1}=N_{t}e^{r\left(1-{\frac {N_{t}}{K}}\right)}
$$

where $r = ln(\lambda)$ and $K$ is the carrying capacity


## Hacking a Ricker model GLM

This is the final model:

```{r}
m_rick <- glm(num.wolves ~ num.prev, offset = log(num.prev),
              family = poisson, data = wolves2)
```



Here's how we got there:

$$
log(N_{t+1})=log(N_{t}) + log\left( (e^r)^{\left(1-{\frac {N_{t}}{K}}\right)}\right)
$$

$$
log(N_{t+1})=log(N_{t}) + log (e^r)\times{\left(1-{\frac {N_{t}}{K}}\right)}
$$

$$
log(N_{t+1})= r-\frac{r}{K}N_{t} + \textrm{offset}[log(N_{t})]
$$


$$
ln(N_{t+1}) = \beta_0 + \beta_1x_1
$$

$\beta_0 = r$

```{r}
coef(m_rick)[1]
```

$\lambda = exp(\beta_0)$

```{r}
exp(coef(m_rick)[1])
```

$\beta_1 = -r/K$
$K = -\beta_0/\beta_1$

```{r}
-coef(m_rick)[1]/coef(m_rick)[2]
```


```{r}
ggplot(wolves2, aes(x = year, y = num.wolves)) +
  geom_point() +
  geom_line(aes(y = predict(m_process, type = "response")), color = "orange") +
  geom_line(aes(y = predict(m_rick, type = "response")), color = "brown")
```

## Quadratic model

A quadratic relationship between $ln(N_T)$ and $T$ with a log-link and observation error approximates the Ricker model

$$
log(N_T) = \beta_0 + \beta_1T + \beta_2T^2
$$

```{r}
m_quad <- glm(num.wolves ~ year_post + I(year_post^2), 
              family = poisson, data = wolves2)
coef(m_quad)
```

NOTE: `I()` = "inhibit interpretation" = "literally do the math, don't mistake this for glm code!"

```{r echo=FALSE}
wolfplot2 +
  geom_line(aes(y = predict(m_quad, type = "response")), color = "purple")
```


## Model competition

```{r}
nrow(wolves) == nrow(wolves2)
```

That's no good because we need to all models to the *same* data.

Re-fit exponential growth model using `wolves2` (alternatively, know what wolf population was in 1994).

```{r}
m_exp2 <- glm(num.wolves ~ year_post, family = poisson, data = wolves2)
```

```{r}
AIC(m_exp2, m_process, m_rick, m_quad)
```


