---
title: "Homework Title"
author: "Your Name"
date: "2/20/2020"
output: word_document
---

**Delete these instructions before you hand in your homework!!**

Use this template for your homeworks, if you want.  It's still totally fine to just type your answers in Word and copy and paste in R code output and plots.

Make sure to edit the YAML header above this text.

TIP: Regularly choose **"restart R and run all chunks"** from the Session menu (or the run button).  If you get errors, then this will NOT knit to word!!

If you don't have Word, then you can knit to PDF or HTML and print the results.

# Common problems using R Markdown:

1. My code runs but it doesn't knit!

  - Did you try "restart R and run all chunks?"  Your code might not actually run in a fresh R session.

  - Make sure your .RMD file name doesn't contain spaces.  Sometimes (but not always) this prevents files from knitting to word correctly

2. Things aren't formatted like they should be in the Word document.

  - Markdown (this space here outside of code chunks) can be finicky.  Just edit the Word doc by hand if you can't figure out how to format something right.

  - Try adding spaces.  Seriously, markdown sometimes gets tripped up when there isn't enough space between, say, text and a code chunk, or a header and a bulleted list.

# Markdown cheatsheet:

# Heading 1
## Heading 2
### Heading 3

*itallics*
**bold**

- bulleted
- list

**Delete these instructions before you hand in your homework!!**
 

```{r packages, include=FALSE}
library(tidyverse)
# load any other packages you need in this chunk
```

```{r data, include=FALSE}
# load all data (.csv files) in this chunk.
# birds <- read_csv("birds.csv")
```


# Question 1

## Part A

```{r}
#your code here
```

<Your written answers here.  Homeworks that only show code and code output will lose points!>

## Part B

```{r}
#your code here
```

<your written answers go here>

# Question 2

## Part A

```{r echo=FALSE}
#This chunk has options set to only show the output, but not the code that generates the output (echo=FALSE)
```



## Part B

```{r}
#code goes here
```

