---
title: "BIO 133 Homework 2 Solutions"
author: "Eric Scott"
date: '2020-02-05'
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# General feedback

### rounding

- don't round until the end 
- should feel comfortable using R as a calculator

### likelihood

- P(data|model) = L(model|data) 
- For a binomial process, this is the output of `dbinom()` OR `logLik(glm(cbind(success, failures) ~ 1, family = binomial(link = "identity)))`
- Bernouli probablities are not the same as binomial probabilities because Bernouli is events in a particular order!  Remember that ${N \choose k}$ is in the binomial probability equation!
- Likelihood ratio and posterior probability ratio should be the same when priors are uninformative.



# 1 Probability for two models

```{r}
p <- c("fair" = 0.5, "weighted" = 0.55)
k <- 110
N <- 200

binProb <- dbinom(k, N, p)
binProb
```

# 2 Calculate posteriors with flat priors

```{r}
priors <- c(0.5, 0.5)
pdata <- sum(binProb * priors)
pmodels <- binProb * priors / pdata
pmodels
```

# 3 Likelihood ratio

```{r}
binProb[2] / binProb[1]

pmodels[2] / pmodels[1]
```

Model 2 is about 2.7x more likely than model 1

OR

```{r}
binProb[1] / binProb[2]
```

LR and probability ratio are the same. 

# 4 What would Nate's prior need to be to make model 1 more likely?

$$
P(model_1|data) = \frac{P(data|model_1)P(model_1)}{P(data)}
$$

$$
P(model_1|data) = \frac{P(data|model_1)P(model_1)}{P(data|model_1)P(model_1) + P(data|model_2)P(model_2)}
$$

1. Set P(model1) to x and P(model2) to 1-x
2. Set P(model1 | data) to 0.5
3. Remember that $P(data) = P(data|model_1)x + P(data|model_2)(1-x)$
3. Solve for x


Nate believes the P(heads) = 0.5

set Nate's prior belief to x

P(model1) = x

P(model2) = 1-x because we are assuming there are only 2 possible models

P(data|model1) = 0.02079869
P(data|model2) = 0.05663140

P(data) = 0.02079869x + 0.05663140(1-x)

0.5 = 0.02079869x/(0.02079869x + 0.05663140(1-x))  

0.5 = 0.02079869x/(0.02079869x + 0.05663140 - 0.05663140x)  

0.5 = 0.02079869x/0.05663140 - 0.03583271x  

0.5(0.05663140 - 0.03583271x) = 0.02079869x  

0.0283157 - 0.01791635x = 0.02079869x  

0.0283157 = 0.03871504x  

x = 0.7313876


A numerical solution using a for-loop (for example) would also have been OK
